A multi-dimensional extension of the Lightweight Temporal Compression method

### 2018-10-17 paper.tex

@@

\begin{abstract}
Lightweight Temporal Compression
(LTC) is among the lossy stream
compression methods that provide the highest compression rate for the
lowest resource (CPU, RAM) consumption. As such, it makes it a very
good candidate for the compression of data streams acquired in
energy-constrained systems such as systems on module (SoMs). The type of data acquired on
such systems, however, is often multi-dimensional. For instance 
accelerometers and gyroscopes usually measure variables along 3 
directions, these variables being distinct but related through the 
actual movement of the sensor. In this paper, we investigate the 
extension of LTC to higher dimensions. First, we provide a formulation 
of the algorithm in an arbitrary vectorial space of dimension $n$. 
Then, we implement the algorithm for the infinity and Euclidean norms, 
in spaces of dimension 2D+t and 3D+t. We evaluate our implementation on 
3D acceleration streams acquired by a SoM during human activities. 
Results show that the 3D implementation of LTC can save up to 20\% in 
energy consumption for low-paced activities such as walking. The 
implementation with the infinity norm only requires about 100~B of 
memory.
\end{abstract}

@@

% Compression of stream data
% need references
Lossy compression of data streams is required in many contexts, to reduce 
storage and network requirements of data analytics. In particular, it 
is highly beneficial to energy-constrained devices such as Systems on 
Modules (SoM), as they reduce the amount of energy-consuming radio 
transmissions. However, resource-intensive algorithms such as the ones 
based on polynomial interpolation or auto-regression 
methods~\cite{lu2010optimized} are not well-suited for 
energy-constrained systems, due to the limited memory available on 
these systems (typically a few KB), and the energy consumption 
associated with CPU usage. On such systems, compression algorithms need 
to find a trade-off between reducing network communications and 
increasing memory and CPU usage. As 
discussed in~\cite{zordan2014performance}, linear compression methods 
provide a very good compromise between these two factors, leading to 
substantial energy reduction.

@@

% LTC summary
% define transmitted, received points, copmression ratio
The Lightweight Temporal Compression method 
(LTC~\cite{schoellhammer2004lightweight}) has been designed 
specifically for energy-constrained systems, initially sensor networks. 
It approximates data points by a piecewise linear function that 
guarantees an upper bound on the reconstruction error. LTC is a very 
lightweight method, with memory requirements in $\mathcal{O}(1)$. 
However, LTC has only been described for 1D streams, which makes it 
limited for multi-dimensional streams such as the ones coming from 
accelerometers, gyroscopes and other movement sensors.

@@

We assume that the stream consists of a sequence of data points 
received at uneven intervals. The compression algorithm 
\emph{transmits} fewer points than it receives. The transmitted points 
might be included in the stream, or computed from stream points. The 
\emph{compression ratio} is the ratio between the number of received 
points and the number of transmitted points. An application would 
reconstruct the stream from the transmitted points: the 
\emph{reconstruction error} is the maximum absolute difference between 
a point of the reconstructed stream, and the corresponding 
point in the original stream.

\begin{figure}
\includegraphics[width=\columnwidth]{ltc.pdf}
\caption{Illustration of the LTC algorithm and notations used. Blue 
dots are received points, red dots are transmitted points. Dashed lines 
represent the high line and the low line at times when a point is 
transmitted.\vspace*{-0.3cm}}
\label{fig:ltc}
\end{figure}

% Outline
Section~\ref{sec:ltc} provides some background on the LTC algorithm, 
and formalizes the description initially proposed 
in~\cite{schoellhammer2004lightweight}. Section~\ref{sec:extension} 
presents our norm-independent extension to dimension $n$, and 
Section~\ref{sec:implementation} describes our implementation. 
Section~\ref{sec:results} reports on experiments to validate our 
implementation, and evaluate the impact of n-dimensional LTC on SoM 
energy consumption.

@@

LTC approximates the data stream
by a piece-wise linear function of time, ensuring that the error at
every time-point is bounded.

@@

The algorithm \emph{receives} a stream of \emph{data points} $x_i$
at times $t_i$ ($i \in \mathbb{N}$), and it \emph{transmits} a stream of data points $\xi_i$
at times $\tau_i$ ($i \in \mathbb{N}$). To simplify the notations, we assume that:
\begin{equation*}
\forall k \in \mathbb{N}, \  \exists ! i \in \mathbb{N} \  \tau_k = t_i
\end{equation*}
That is, transmission times coincide with reception times.
We define the \emph{centered received points} as follows:
\begin{equation*}
\forall k \in \mathbb{N}\ , \forall j \in \mathbb{N^*},\ (u^k_j, y^k_j) = (t_{i+j}, x_{i+j}), \quad \mathrm{where\ }i\mathrm{\ is\ s.t.}\ t_i = \tau_k.
\end{equation*}
and:
\begin{equation*}
\forall k \in \mathbb{N},\  (u^k_0, y^k_0) = (\tau_k, \xi_k).
\end{equation*}
This definition is such that $y^k_j$ is the $j^{th}$ data point received
after the $k^{th}$ transmission.
Figure~\ref{fig:ltc} illustrates the notations and algorithm.

@@

\begin{algorithm}
\begin{algorithmic}[1]
\State tr = $(u^0_0, y^0_0)$ \Comment{last transmitted point}
\State k = 0 ; j = 1
\State (lp, hp) = ($y^1_1 - \epsilon$, $y^1_1 + \epsilon$) \Comment{low and high points}

\While{True} \Comment{Process received points as they come}
    \State j += 1
    \State new\_lp = max($y^k_j-\epsilon$, line($u^k_j$, tr, ($u^k_{j-1}$, lp)))
    \State new\_hp = max($y^k_j+\epsilon$, line($u^k_j$, tr, ($u^k_{j-1}$, hp)))
    \If{new\_lp $<$ new\_hp} \Comment{Keep compressing}
        \State (lp = new\_lp, new\_hp)
        \State \textbf{continue}
    \EndIf
    \State tr = $(u^k_j, (lp+hp)/2)$
    \Comment{Transmit point}
    \State (lp, hp) = ($y^k_j-\epsilon$, $y^k_j+\epsilon$)
    \State k += 1
    \State j = 0
\EndWhile
\end{algorithmic}
\caption{Original LTC algorithm, adapted from~\cite{schoellhammer2004lightweight}.}
\label{algo:ltc}
\end{algorithm}

@@

Let $(u_0^k, y_0^k) \in \mathbb{R}^{n+1}$ be the latest transmitted point. For convenience, all the subsequent points will be
expressed in the orthogonal space with origin $(\tau^k, \xi^k)$. We denote by $(v_j, z_j)_{j \in \llbracket 0, m \rrbracket}$ such points:
\begin{equation*}
\forall j \leq m,\  (v_j, z_j) = (u_j^k - \tau^k, y_j^k - \xi^k)
\end{equation*}
We define $\mathcal{B}_j$ as the ball of $\mathbb{R}^n$ of center $\frac{1}{j}z_j$ and of radius
$\frac{\epsilon}{j}$:
\begin{equation*}
\mathcal{B}_j = \left\{ z \in \mathbb{R}^n,\  \norm{z-\frac{v_1}{v_j}z_j} \leq \frac{v_1}{v_j}\epsilon \right\}
\end{equation*}

\subsection{LTC property}

We formulate the \emph{LTC property} as follows:
\begin{equation*}
\exists z \in \mathbb{R}^n, \ \forall j \in \llbracket 1, m \rrbracket, \norm{\frac{v_j}{v_1}z-z_j} \leq \epsilon.
\end{equation*}
The LTC algorithm ensures that the LTC property is
verified between each transmission. Indeed, all the data points
$z$ such that $(v_1, z)$ is between the high line and the low line
verify the property. Line 9 in Algorithm~\ref{algo:ltc} guarantees that
such a point exists.
The LTC property can be written as follows:
\begin{equation*}
\exists z \in \mathbb{R}^n, \ \forall j \in \llbracket 1, m \rrbracket, \norm{z-\frac{v_1}{v_j}z_j} \leq \frac{v_1}{v_j}\epsilon
\end{equation*}
that is:
\begin{equation}
\bigcap_{j=1}^m \mathcal{B}_j \neq \O
\label{eq:ltc-property}
\end{equation}
Note that $(\mathcal{B}_j)_{j \in \llbracket 1, m \rrbracket}$ is a sequence
of balls of strictly decreasing radius, since $v_j > v_1$.
\todo{Add a figure to illustrate this}

@@

The LTC algorithm generalized to dimension $n$ test, for each data point,
that the LTC property in Equation~\ref{eq:ltc-property} is verified. It is written in
Algorithm~\ref{algo:general-ltc}.
\begin{algorithm}
\begin{algorithmic}[1]
\State tr = ($\tau, \xi$) = ($u^0_0, y^0_0$)
\State k = 0 ; j = 0
\While{True}
    \State j += 1
    \State ($v_j, z_j$) = ($u_j^k - \tau, y_j^k - \xi$)
    \If{$\bigcap_{l=1}^j{\mathcal{B}_l} \neq \O$}
        \State \textbf{continue}
    \EndIf
    \State Pick $z$ in $\bigcap_{l=1}^{j-1}{\mathcal{B}_j}$
    \State tr = ($\tau$, $\xi$) = ($u^k_{j-1}, z$)
    \State k += 1
    \State j = 0
\EndWhile
\end{algorithmic}
\caption{Generalized LTC}
\label{algo:general-ltc}
\end{algorithm}

@@

Although Algorithm~\ref{algo:general-ltc} looks simple, one should not
overlook the fact that there is no good general algorithm to test
whether a set of balls intersect. The best general algorithm we could find
so far relies on Helly's theorem which is formulated as follows~\cite{helly1923mengen}:
\begin{theorem}
Let $\left\{ X_i \right\}_{i \in \llbracket 1, m \rrbracket}$ be a collection of convex subsets of $\mathbb{R}^n$. If the intersection of every $n+1$
subsets is non-empty, then the whole collection has an non-empty intersection.
\end{theorem}
This theorem leads to an algorithm of complexity ${m \choose n+1}$ which is
not usable in resource-constrained environment.

The only feasible algorithm that we found is norm-specific. It
maintains a representation of the intersection
$\bigcap_{j=1}^{m}{\mathcal{B}_m}$ which is updated at every iteration.
The intersection tests can then be done in constant time, however,
updating the representation of the intersection may be more costly
depending on the norm used. For the infinity norm, the representation
is a rectangular cuboid which is straightforward to update by
intersection with an n-ball, since the n-ball is a rectangular cuboid too.
For the Euclidean norm, the representation is a volume with no specific property,
which is more costly to maintain.

@@

As mentioned before, norm selection in $\mathbb{R}^n$ has a critical
impact on the compression error and ratio. To appreciate this effect,
let us consider the comparison between the infinity norm and the
Euclidean norm in dimension 2. By comparing the unit disk to the unit
square, we obtain that the compression ratio of a random stream would
be $\frac{4}{\pi}$ times larger with the infinity norm than with the 
Euclidean norm. In 3D, this ratio would be $\frac{6}{\pi}$. Conversely, 
a compression error bounded by $\epsilon$ with the infinity norm 
corresponds to a compression error of $\frac{\epsilon}{\sqrt{n}}$ with 
the Euclidean norm. Unsurprisingly, the
infinity norm is more 'tolerant' than the Euclidean norm.

\todo{Perhaps add a figure}

It should also be noted that using the infinity norm boils down to the
use of the 1D LTC algorithm independently in each dimension: a data
point is transmitted as soon as the linear approximation doesn't hold
in any of the coordinates. \todo{This is unclear}

@@

@@

We conducted two experiments using Motsai's Neblina 
module\footnote{\url{ https://motsai.com/products/neblina}}, a system 
with a Nordic Semiconductor nRF52832 micro-controller, 64~KB of RAM, 
and Bluetooth Low Energy connectivity. The Neblina has a 3D 
accelerometer, a 3D gyroscope, a 3D magnetometer, and environmental sensors 
for humidity, temperature and pressure. It can function as an Inertial 
Motion Unit, a Vertical Reference Unit, or an Attitude and Heading 
Reference System. The Neblina has a battery of 100mAh; its average 
consumption is 2.52~mA without radio transmission and 3.47~mA with 
radio transmission, leading to an autonomy of 39.7~h without 
transmission and 28.8~h with transmission. \todo{Omid, could you review
this paragraph?}

\subsection{Experiment 1: validation tests}

In this experiment we validated the behavior of our LTC extensions on a 
PC processing data acquired with the Neblina. We collected two 3D 
accelerometer time series, a short one and a longer one, acquired 
on two different subjects performing biceps curl, with a 50Hz sampling rate (see 
Figure~\ref{fig:datasets-1}). It should be noted that the longer dataset also has 
a higher amplitude, most likely due to differences between subjects.

\begin{figure*}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.33\columnwidth]{figures/5-time-bicep-curl-plot-x.pdf}
\includegraphics[width=0.33\columnwidth]{figures/5-time-bicep-curl-plot-y.pdf}
\includegraphics[width=0.33\columnwidth]{figures/5-time-bicep-curl-plot-z.pdf}
\caption{Short}
\end{subfigure}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.33\columnwidth]{figures/Mohammad-bicep-curl-plot-x.pdf}
\includegraphics[width=0.33\columnwidth]{figures/Mohammad-bicep-curl-plot-y.pdf}
\includegraphics[width=0.33\columnwidth]{figures/Mohammad-bicep-curl-plot-z.pdf}
\caption{Long} 
\end{subfigure}
\caption{Time series used in Experiment 1}
\label{fig:datasets-1}
\end{figure*}

We compressed the time series with various values of $\epsilon$, using our 
2D (x and y) and 3D (x, y and z) implementations of LTC. On the Neblina,
the resolution of the accelerometer is about 20~mg (196~mm/$s^2$).
 We used a 
laptop computer with 16~GB of memory, and Intel i5-3210M CPU @ 2.50GHz 
× 4, and Linux Fedora 27. We measured memory consumption using 
Valgrind's massif 
tool~\cite{nethercote2006building}\footnote{\url{http://valgrind.org}}, 
and processing time using \texttt{gettimeofday()} from the GNU C 
Library.

@@

\paragraph{Inifinity vs Euclidean norms}
Compared to the infinity norm, The average ratio between the compression ratios obtained
with the infinity and Euclidean norm is 1.08 for 2D data, and 1.20 
for 3D data. These ratios are lower than the theoretical values of 
$\frac{4}{\pi}$ in 2D and $\frac{6}{\pi}$ in 3D, which are obtained for 
random-uniform signals. Unsurprisingly, the infinity norm surpasses the 
Euclidean norm in terms of resource consumption. Memory-wise, the 
infinity norm requires a constant amount of 80~B, used to store the 
intersection of n-balls. This amount is realistic on modules such as
the Neblina. The Euclidean norm, however, uses up to 4.9~KB of memory 
on the Long time series in 3D with $\epsilon=100$. More importantly, 
the amount of required memory increases for more complex time series 
(Long vs Short), and it also increases with larger 
values of $\epsilon$. Similar observations are made for the processing 
time, with values ranging from 0.2~ms for the simplest time series and 
smallest $\epsilon$, to 26.4~ms for the most complex time series and 
largest $\epsilon$. A better algorithm needs to be found to make the 
Euclidean norm usable for systems on modules.

\paragraph{2D vs 3D}
For a given $\epsilon$ of 100 and a given time series, the compression 
ratios are always higher in 2D than in 3D. It makes sense since the 
\todo{x} is higher in 3D than it is in 2D. Resource consumption is 
higher in 3D than in 2D: for the infinity norm, 3D consumes 1.4 times 
more memory than 2D (1.6 times on average for Euclidean norm), and 
processing time is 1.3 longer on average (1.18 on average for Euclidean 
norm).

>>>>

\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}

@@

\begin{abstract}
Lightweight Temporal Compression
(LTC) is among the lossy stream
compression methods that provide the highest compression rate for the 
lowest CPU and memory consumption. As such, it is very well suited to 
compress data streams in energy-constrained systems such as connected 
objects. The current formulation of LTC, however, is one-dimensional 
while data acquired in connected objects is often multi-dimensional: for instance, 
accelerometers and gyroscopes usually measure variables along 
3 directions related by the actual movement of the sensor. In this 
paper, we investigate the extension of LTC to 
higher dimensions. First, we provide a formulation 
of the algorithm in an arbitrary vectorial space of dimension $n$. 
Then, we implement the algorithm for the infinity and Euclidean norms, 
in spaces of dimension 2D+t and 3D+t. We evaluate our implementation on 
3D acceleration streams acquired during human activities. 
Results show that the 3D implementation of LTC can save up to 20\% in 
energy consumption for low-paced activities, with a memory usage of about 100~B.
\end{abstract}

@@

With the recent technological advances in Internet of Things (IoT) 
applications, more than one billion connected objects are expected to 
be launched worldwide by 2025\footnote{\url{ 
https://www.statista.com/statistics/471264/iot-number-of-connected-devices-worldwide}} 
Power consumption is amongst the biggest challenges targeting connected 
objects, particularly in the industrial domains, where several sensing 
systems are commonly launched in the field to run for days or even 
weeks without being recharged. Typically, such devices use sensors to 
capture properties such as temperature or motion, and stream them to a 
host system over a radio transmission protocol such as Bluetooth 
Low-Energy (BLE). System designers aim to reduce the rate of data 
transmission as much as possible, as radio transmission is a 
power-hungry operation.

 Compression is a key technique to reduce the rate of radio 
 transmission.  While in several applications lossless compression 
 methods are more desirable than lossy compression techniques, in the 
 context of IoT and sensor data streams, the measured sensor data 
 intrinsically involves noise and measurement errors, which can 
 be treated as a configurable 
tolerance for a lossy compression algorithm.

Resource-intensive lossy compression algorithms such as the ones based on 
polynomial interpolation, discrete cosine and Fourier transforms, or 
auto-regression methods~\cite{lu2010optimized} are not well-suited for 
energy-constrained systems, due to the limited memory available on 
these systems (typically a few KB), and the energy consumption 
associated with CPU usage. On such systems, compression algorithms need 
to find a trade-off between reducing network communications and 
increasing memory and CPU usage. As 
discussed in~\cite{zordan2014performance}, linear compression methods 
provide a very good compromise between these two factors, leading to 
substantial energy reduction.

@@

% LTC summary
% define transmitted, received points, copmression ratio
The Lightweight Temporal Compression method 
(LTC~\cite{schoellhammer2004lightweight}) has been designed 
specifically for energy-constrained systems, initially sensor networks. 
It approximates data points by a piecewise linear function that 
guarantees an upper bound on the reconstruction error, and a reduced 
memory footprint in $\mathcal{O}(1)$. However, LTC has only been 
described for 1D streams, while streams acquired by connected objects, such as 
acceleration or gyroscopic data, are often multi-dimensional.

@@

We assume that the stream consists of a sequence of data points 
received at uneven intervals. The compression algorithm 
\emph{transmits} fewer points than it receives. The transmitted points 
might be included in the stream, or computed from stream points. The 
\emph{compression ratio} is the ratio between the number of received 
points and the number of transmitted points. An application 
reconstructs the stream from the transmitted points: the 
\emph{reconstruction error} is the maximum absolute difference between 
a point of the reconstructed stream, and the corresponding 
point in the original stream.

\begin{figure}
\includegraphics[width=\columnwidth]{ltc.pdf}
\caption{Illustration of the LTC algorithm and notations used. Blue 
dots are received points, red dots are transmitted points. Dashed lines 
represent the high line and the low line at times when a point is 
transmitted.\vspace*{-0.3cm}}
\label{fig:ltc}
\end{figure}
\todo{show $\xi$ in the Figure}

% Outline
Section~\ref{sec:ltc} provides some background on the LTC algorithm, 
and formalizes the description initially proposed 
in~\cite{schoellhammer2004lightweight}. Section~\ref{sec:extension} 
presents our norm-independent extension to dimension $n$, and 
Section~\ref{sec:implementation} describes our implementation. 
Section~\ref{sec:results} reports on experiments to validate our 
implementation, and evaluates the impact of n-dimensional LTC on 
energy consumption of connected objects.

@@

LTC approximates the data stream
by a piece-wise linear function of time, ensuring that the error at
every time-point is bounded by parameter $\epsilon$.

@@

The algorithm receives a stream of data points $x_i$
at times $t_i$ ($i \in \mathbb{N}$), and it transmits a stream of data points $\xi_i$
at times $\tau_i$ ($i \in \mathbb{N}$). To simplify the notations, we assume that:
\begin{equation*}
\forall k \in \mathbb{N}, \  \exists ! i \in \mathbb{N} \  \tau_k = t_i
\end{equation*}
That is, transmission times coincide with reception times.
We define the \emph{shifted received points} as follows:
\begin{equation*}
\forall k \in \mathbb{N}\ , \forall j \in \mathbb{N^*},\ (u^k_j, y^k_j) = (t_{i+j}, x_{i+j}), 
\end{equation*}
where $i$ is such that $t_i = \tau_k$ and:
\begin{equation*}
\forall k \in \mathbb{N},\  (u^k_0, y^k_0) = (\tau_k, \xi_k).
\end{equation*}
This definition is such that $y^k_j$ is the $j^{th}$ data point received
after the $k^{th}$ transmission and $u^k_j$ is the corresponding timestamp.
Figure~\ref{fig:ltc} illustrates the notations and algorithm.

@@

\begin{algorithm}
\begin{algorithmic}[1]
\Input
   \Desc{$(u^k_j, y^k_j)$}{$\quad \quad $Received data stream}
   \Desc{$\epsilon$}{$\quad \quad$Error bound}
\EndInput
\Output
   \Desc{tr}{Transmitted points}
\EndOutput
\State tr = $(u^0_0, y^0_0)$ \Comment{last transmitted point}
\State k = 0 ; j = 1
\State (lp, hp) = ($y^1_1 - \epsilon$, $y^1_1 + \epsilon$) \Comment{low and high points}

\While{True} \Comment{Process received points as they come}
    \State j += 1
    \State new\_lp = max($y^k_j-\epsilon$, line($u^k_j$, tr, ($u^k_{j-1}$, lp)))
    \State new\_hp = max($y^k_j+\epsilon$, line($u^k_j$, tr, ($u^k_{j-1}$, hp)))
    \If{new\_lp $<$ new\_hp} \Comment{Keep compressing}
        \State (lp, hp) = (new\_lp, new\_hp)
        \State \textbf{continue}
    \EndIf
    \State tr = $(u^k_j, (lp+hp)/2)$
    \Comment{Transmit point}
    \State (lp, hp) = ($y^k_j-\epsilon$, $y^k_j+\epsilon$)
    \State k += 1
    \State j = 0
\EndWhile
\end{algorithmic}
\caption{Original LTC algorithm, adapted from~\cite{schoellhammer2004lightweight}.}
\label{algo:ltc}
\end{algorithm}

@@

Let $(u_0^k, y_0^k) \in \mathbb{R}^{n+1}$ be the latest transmitted point. For convenience, all the subsequent points will be
expressed in the orthogonal space with origin $(\tau_k, \xi_k)$. We denote by $(v_j, z_j)_{j \in \llbracket 0, m \rrbracket}$ such points:
\begin{equation*}
\forall j \leq m,\  (v_j, z_j) = (u_j^k - \tau_k, y_j^k - \xi_k)
\end{equation*}
We define $\mathcal{B}_j$ as the ball of $\mathbb{R}^n$ of center $\frac{v_1}{v_j}z_j$ and of radius
$\frac{v_1}{v_j}\epsilon$:
\begin{equation*}
\mathcal{B}_j = \left\{ z \in \mathbb{R}^n,\  \norm{z-\frac{v_1}{v_j}z_j} \leq \frac{v_1}{v_j}\epsilon \right\}
\end{equation*}

\subsubsection{LTC property}

We formulate the \emph{LTC property} as follows:
\begin{equation*}
\exists z \in \mathbb{R}^n, \ \forall j \in \llbracket 1, m \rrbracket, \norm{\frac{v_j}{v_1}z-z_j} \leq \epsilon.
\end{equation*}
The original LTC algorithm ensures that the LTC property is
verified between each transmission. Indeed, all the data points
$z$ such that $(v_1, z)$ is between the high line and the low line
verify the property. Line 9 in Algorithm~\ref{algo:ltc} guarantees that
such a point exists.

The LTC property can be re-written as follows:
\begin{equation*}
\exists z \in \mathbb{R}^n, \ \forall j \in \llbracket 1, m \rrbracket, \norm{z-\frac{v_1}{v_j}z_j} \leq \frac{v_1}{v_j}\epsilon
\end{equation*}
that is:
\begin{equation}
\bigcap_{j=1}^m \mathcal{B}_j \neq \O
\label{eq:ltc-property}
\end{equation}
Note that $(\mathcal{B}_j)_{j \in \llbracket 1, m \rrbracket}$ is a sequence
of balls of strictly decreasing radius, since $v_j > v_1$.

@@

The LTC algorithm generalized to dimension $n$ tests that the LTC 
property in Equation~\ref{eq:ltc-property} is verified after each reception of a data 
point. It is written in Algorithm~\ref{algo:general-ltc}.
\begin{algorithm}
\begin{algorithmic}[1]
\Input
   \Desc{$(u^k_j, y^k_j)$}{$\quad \quad $Received data stream}
   \Desc{$\epsilon$}{$\quad \quad$Error bound}
\EndInput
\Output
   \Desc{tr}{Transmitted points}
\EndOutput

\State tr = ($\tau, \xi$) = ($u^0_0, y^0_0$)
\State k = 0 ; j = 0
\While{True}
    \State j += 1
    \State ($v_j, z_j$) = ($u_j^k - \tau, y_j^k - \xi$)
    \If{$\bigcap_{l=1}^j{\mathcal{B}_l} \neq \O$}
        \State \textbf{continue}
    \EndIf
    \State Pick $z$ in $\bigcap_{l=1}^{j-1}{\mathcal{B}_j}$
    \State tr = ($\tau$, $\xi$) = ($u^k_{j-1}, z$)
    \State k += 1
    \State j = 0
\EndWhile
\end{algorithmic}
\caption{Generalized LTC}
\label{algo:general-ltc}
\end{algorithm}

@@

Although Algorithm~\ref{algo:general-ltc} looks simple, one should not
overlook the fact that there is no good general algorithm to test
whether a set of balls intersect. The best general algorithm we could find
so far relies on Helly's theorem which is formulated as follows~\cite{helly1923mengen}:
\begin{theorem}
Let $\left\{ X_i \right\}_{i \in \llbracket 1, m \rrbracket}$ be a collection of convex subsets of $\mathbb{R}^n$. If the intersection of every $n+1$
subsets is non-empty, then the whole collection has an non-empty intersection.
\end{theorem}
\noindent This theorem leads to an algorithm of complexity ${m \choose n+1}$ which is
not usable in resource-constrained environment.

The only feasible algorithm that we found is norm-specific. It
maintains a representation of the intersection
$\bigcap_{j=1}^{m}{\mathcal{B}_j}$ which is updated at every iteration.
The intersection tests can then be done in constant time. However,
updating the representation of the intersection may be costly
depending on the norm used. For the infinity norm, the representation
is a rectangular cuboid which is straightforward to update by
intersection with an n-ball, since the n-ball is a rectangular cuboid too.
For the Euclidean norm, the representation is a volume with no particular property,
which is more costly to maintain.

@@

As mentioned before, norm selection in $\mathbb{R}^n$ has a critical
impact on the compression error and ratio. To appreciate this effect,
let us consider the comparison between the infinity norm and the
Euclidean norm in dimension 2. By comparing the unit disk to a
square of side 2, we obtain that the compression ratio of a random stream would
be $\frac{4}{\pi}$ times larger with the infinity norm than with the 
Euclidean norm. In 3D, this ratio would be $\frac{6}{\pi}$. Conversely, 
a compression error bounded by $\epsilon$ with the infinity norm 
corresponds to a compression error of $\frac{\epsilon}{\sqrt{n}}$ with 
the Euclidean norm. Unsurprisingly, the
infinity norm is more 'tolerant' than the Euclidean norm.

It should also be noted that using the infinity norm in dimension $n$ 
boils down to the use of the 1D LTC algorithm independently in each 
dimension, since a data point will be transmitted as soon as the linear 
approximation doesn't hold in any of the dimensions.

@@

\todo{explain difference between exp 1 and exp 2 for euclidean}

@@

\todo{talk about data shift}

We conducted two experiments using Motsai's Neblina 
module\footnote{\url{ https://motsai.com/products/neblina}}, a system 
with a Nordic Semiconductor nRF52832 micro-controller, 64~KB of RAM, 
and Bluetooth Low Energy connectivity. Neblina has a 3D 
accelerometer, a 3D gyroscope, a 3D magnetometer, and environmental 
sensors for humidity, temperature and pressure. The platform is 
equipped with sensor fusion algorithms for 3D orientation tracking and 
a machine learning engine for complex motion analysis and motion 
pattern recognition~\cite{sarbishei2016accuracy}. Neblina has a 
battery of 100mAh; its average consumption is 2.52~mA when using 
accelerometer and gyroscope sensors at 200~Hz but without radio 
transmission, and 3.47~mA with radio transmission, leading to an 
autonomy of 39.7~h without transmission and 28.8~h with transmission.

\subsection{Experiment 1: validation}

\todo{In all of our experiments, the board was attached to the lower 
right arm as shown in Fig. X. (Put a picture of Neblina in a box worn 
on the wrist)}

In this experiment, we validated the behavior of our LTC extension on a 
PC using data acquired with Neblina. We collected two 3D 
accelerometer time series, a short one and a longer one, acquired 
on two different subjects performing biceps curl, with a 50Hz sampling rate (see 
Figure~\ref{fig:datasets-1}). It should be noted that the longer dataset also has 
a higher amplitude, most likely due to differences between subjects.

\begin{figure*}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.33\columnwidth]{figures/5-time-bicep-curl-plot-x.pdf}
\includegraphics[width=0.33\columnwidth]{figures/5-time-bicep-curl-plot-y.pdf}
\includegraphics[width=0.33\columnwidth]{figures/5-time-bicep-curl-plot-z.pdf}
\caption{Short biceps curl}
\end{subfigure}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.33\columnwidth]{figures/Mohammad-bicep-curl-plot-x.pdf}
\includegraphics[width=0.33\columnwidth]{figures/Mohammad-bicep-curl-plot-y.pdf}
\includegraphics[width=0.33\columnwidth]{figures/Mohammad-bicep-curl-plot-z.pdf}
\caption{Long biceps curl} 
\end{subfigure}
\caption{Time series used in Experiment 1}
\label{fig:datasets-1}
\end{figure*}

We compressed the time series with various values of $\epsilon$, using 
our 2D (x and y) and 3D (x, y and z) implementations of LTC. On 
Neblina, the raw uncalibrated accelerometer data corresponds to errors 
around 20~mg (196~mm/$s^2$). \todo{Check g units} We used a 
laptop computer with 16~GB of RAM, an Intel i5-3210M CPU @ 2.50GHz 
× 4, and Linux Fedora 27. We measured memory consumption using 
Valgrind's massif 
tool~\cite{nethercote2006building}\footnote{\url{http://valgrind.org}}, 
and processing time using \texttt{gettimeofday()} from the GNU C 
Library.

@@

\paragraph{Inifinity vs Euclidean norms}
Compared to the infinity norm, the average ratio between the compression ratios obtained
with the infinity and Euclidean norm is 1.08 for 2D data, and 1.20 
for 3D data. These ratios are lower than the theoretical values of 
$\frac{4}{\pi}$ in 2D and $\frac{6}{\pi}$ in 3D, which are obtained for 
random-uniform signals. Unsurprisingly, the infinity norm surpasses the 
Euclidean norm in terms of resource consumption. Memory-wise, the 
infinity norm requires a constant amount of 80~B, used to store the 
intersection of n-balls. The Euclidean norm, however, uses up to 4.9~KB of memory 
for the Long time series in 3D with $\epsilon=100$. More importantly, 
the amount of required memory increases for more complex time series 
(Long vs Short), and it also increases with larger 
values of $\epsilon$. Similar observations are made for the processing 
time, with values ranging from 0.2~ms for the simplest time series and 
smallest $\epsilon$, to 26.4~ms for the most complex time series and 
largest $\epsilon$.

\paragraph{2D vs 3D}
For a given $\epsilon$ of 100 and a given time series, the compression 
ratios are always higher in 2D than in 3D. It makes sense since the 
probability for the signal to deviate from a straight line 
approximation is higher in 3D than it is in 2D. Besides, resource 
consumption is higher in 3D than in 2D: for the infinity norm, 3D 
consumes 1.4 times more memory than 2D (1.6 times on average for 
Euclidean norm), and processing time is 1.3 longer on average (1.18 on 
average for Euclidean norm).
